---
title: "Challenge 1"
author: "Audrey Bertin"
description: "Reading in data and creating a post"
date: "5/30/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - faostat
  - wildbirds
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)
library(readxl)
knitr::opts_chunk$set(echo = TRUE)
```

For this challenge, I'll be reading in the following datasets:

-   birds.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐


First, we start with `birds.csv`:

```{r}
birds <- readr::read_csv("_data/birds.csv")
```

A sample of this data can be seen below:

```{r}
head(birds)
```

```{r}
str(birds)
```


As we can see, we appear to have information on the number of different types of birds in different areas around the world, based on different data sources and estimates. The data appears to be measuring birds in captivity, rather than wild birds.

We have 8 categorical variables and 6 numeric. The numeric variables appear to be discrete in that they can only have whole integer values.

The different types of birds we are tracking are as follows. We have the most data on chickens and the least on pigeons and other birds.

```{r}
table(birds$Item)
```

The different type of measurements we have are:

```{r}
table(birds$`Flag Description`)
```
There appears to be a strange duplicate variable. Both `Year` and `Year Code` have identical values.


The birds data has 30,977 rows and 14 columns:

```{r}
dim(birds)
```

Looking at missing data, most of our variables have records for every row. There are only two that have missing values, `Value` and `Flag`. About 1/30 of the Values are missing and about 1/3 of the flags.

```{r}
colSums(is.na(birds))
```
