---
title: "Final Project -- Formula 1 Rule Change Analysis"
author: "Audrey Bertin"
description: "How did the 2022 F1 Rule Change Affect the Sport?"
date: "7/10/2023"
format:
  html:
    df-print: paged
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - final_project
  - formula_1
  - audrey_bertin

---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

I have been regularly watching Formula 1 for the last few years. In that time, there was a significant rule change that--purely observationally--appears to have changed the sport. It is not clear entirely how much or in what ways, so in this project, I want to dig into the data to see if I can come to some evidence-based conclusions.

## F1: A Primer

Formula 1 (F1) is the top class of single-seater motorsport racing. "Formula" refers to the set of rules to which all participants' cars must conform. For example, Formula 2 has different rules about car design and racing than Formula 1.

An F1 season consists of a series of races, known as Grands Prix (GP), which take place around the world on purpose-built racing circuits as well as public roads (termed "street circuits"). The number and location of races per season has varied significantly across the years.

There are currently 10 teams in Formula 1, officially termed "Constructors". Like the races, the number of teams (and names) has varied significantly over time. Each team has *two* drivers who race for them each season (as well as a driver on standby in case one of the primary drivers get sick/injured). 

Each Constructor is responsible for designing their own car each season within a shared set of regulations that restrict the design across all teams (e.g. a minimum weight, certain aerodynamic requirements, etc.). Some Constructors are known as "works teams". These teams design and construct the engines for their cars *in addition* to the aerodynamic components. The rest of the teams are called "customer teams". They design the aerodynamic parts of their car, but *purchase* their engines from one of the works teams instead of designing it themselves. For example, currently, the Mercedes *works* team uses their engines for their own car and  *also* provides engines to the Williams and McLaren teams. 

There is also the concept of "junior teams", where one company essentially has ownership over multiple teams in F1. They have one primary, A-team, and then a B-team that they can use to develop talent for their A-team. For example, currently Red Bull owns both the Red Bull Racing *and* Alpha Tauri teams, and regularly promote their drivers from Alpha Tauri to Red Bull Racing. For the purposes of the championship, however, these are considered *separate* constructors.

Because each Constructor designs their own car, cars can vary quite dramatically across teams, so there tends to be a somewhat clear hierarchy. The "worst" car often has nearly zero chance of winning a race, and the "best" cars almost always end up on the podium. In formula 1 lingo, teams are broken into three categories:

- "top" teams -- these teams win most of the races / usually make up the podium
- "midfield" teams -- these teams are not as fast as the top teams but often still place reasonably well and can, on occasion, end up on the podium or even (in rare cases) win a race
- "backmarker" teams -- these teams are significantly slower than the others, with *almost no* chance of ending up on the podium, and usually finish near the back of the pack

Due to this team hierarchy, it can be difficult to compare a driver from one team to a driver on another. Just because one driver always has faster lap times or finishes in a better position does not mean that they are necessarily a better driver, and could just mean that they have a better car. Since both teammates on a given team drive the same car, driver skill is often judged more-so based on how well a driver does compared to their teammates. 

In each season, there are two separate championships up for grabs -- a **Drivers Championship** and a **Constructors Championship**. The Drivers Championship is considered more prestigious and awards the individual driver who received the most points throughout the season. The Constructors Championship is awarded to the *team* with the most *combined* points across their two drivers. 

Teams and drivers have the chance to earn points at each Grand Prix (note that there are also points available in what are called "Sprint races", but these are a new feature in F1 and we'll be excluding them for the purposes of this analysis in order to simplify things).

Each GP occurs over one weekend is broken up into two segments: qualifying (typically on Saturday) and the race (typically on Sunday). In qualifying, drivers try to put together the *fastest* lap. Lap times then determine the order that the cars will start for the race.

Within the race, cars are required to make *at least one* pit stop to change which type of tires they are using (there are hard, medium, and soft tires available for each race, each of which have different pros and cons). Depending on how much a given track damages the tires, along with other factors such as if a car is involved in a collision, additional pitstops may take place to change tires or swap out parts.

At the end of the race, points are awarded based on drivers' finishing position. Additionally, the driver who records the fastest single lap throughout the race is awarded an extra point in addition to their points for finishing position. 

The current points distribution is as follows:

- 1st place: 25 pts
- 2nd place: 18 pts
- 3rd place: 15 pts
- 4th place: 12 pts
- 5th place: 10 pts
- 6th place: 8 pts
- 7th place: 6 pts
- 8th place: 4 pts
- 9th place: 2 pts
- 10th place: 1 pt
- Fastest Lap: 1 pt 
- 11th-20th place: 0 pts

Points are cumulative across a season, and the championships are awarded to the team and driver with the highest point totals.

## Dataset

Data on Formula 1 is available through Kaggle at the following link
https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020

It contains F1 race data from the series' start in 1950 to the present season, which is currently in-progress and will run until November. This data was likely scraped from the official F1 website, which publishes detailed information after every race, and so it is likely to be accurate and official.

There are 14 separate files within this dataset that store different types of information:

- `circuits.csv` contains information about the different tracks the drivers race on (a row here is a track). There are 77 total circuits raced on since 1950.
- `constructor_standings.csv` contains information about how many points each team has after each race (a row here is a single team at a single race). We have 12,941 data points for the standings.
- `constructors.csv` stores the names and nationalities of teams (a row here is a team). There have been 211 total teams since 1950, and there are currently 10 active teams.
- `driver_standings.csv` stores information about how many points each driver was awarded at each race (a row here is a single driver at a single race). We have 33,902 records on drivers standings across all of the races. 
- `drivers.csv` stores the names and nationalities of drivers (a row here is a driver). There are 857 recorded drivers.
- `lap_times.csv` stores the lap times for every driver on every lap of each race (a row here is a single lap for a single driver at a single race). We have 538,121 lap times.
- `pit_stops.csv` stores information about every pit stop taken during each race (a row here is a single pit stop taken by a single driver at a single race). We have 9,634 recorded pit stops.
- `qualifying.csv` stores qualifying lap times for each driver from each GP (a row here is a single driver in a single GP, and each row contains all qualifying lap times across multiple sessions). We have 9,575 observations here.
- `races.csv` contains information about each GP, e.g. track name, date, etc. (a row here is a single GP). There are 1102 recorded races here.
- `results.csv` contains all of the finishing results for each race, e.g. driver finishing positions, # laps completed, fastest lap (a row here is a single driver at a single race). There are 25,840 observations.
- `seasons.csv` contains the years and urls for wikipedia pages associated with each season (a row here is a season). There are 74 seasons, covering 1950-2023.
- `sprint_results.csv` same as results.csv but for sprint races, which we'll be excluding as they are currently just a new feature that F1 is testing (a row here is a single sprint race). There are 120 recorded observations, a very small number since there have only been a handful of these.s
- `status.csv` contains the encodings of a `status` variable that is referenced in the results files (a row here is a status). There are 139 unique statuses.

We complete a cleaning of these datasets below. See the [associated homework 2 blog post](https://dacss.github.io/DACSS_601_Summer2023_Sec1/posts/AudreyBertin_Homework2.html) for more details on these modifications and what each dataset looked like before and after.

The main goals of these data cleaning steps are to:

1. Remove any unclear variables
2. Get any time related variables into an easily usable format (seconds) and convert any other variables that are not in a correct format
3. Rename variables to avoid duplicates when joining tables and to make each variable's meaning more clear

There is no pivoting involved in this step. Instead, this will be done as needed during the analysis of each research question (based on the question at hand and whether such reshaping is appropriate).


```{r message=FALSE, warning=FALSE}
# Circuits
circuits <- readr::read_csv("_data/formula1/circuits.csv")
circuits <- circuits %>%
  rename(circuit_name = name) %>%
  select(-alt)

# Constructors Standings
constructor_standings <- readr::read_csv("_data/formula1/constructor_standings.csv")
constructor_standings <- constructor_standings %>%
  select(-positionText)

# Constructors
constructors <- readr::read_csv("_data/formula1/constructors.csv")
constructors <- constructors %>%
  rename(constructor_name = name)

# Drivers Standings
driver_standings <- readr::read_csv("_data/formula1/driver_standings.csv")
driver_standings <- driver_standings %>%
  select(- positionText)

# Drivers
drivers <- readr::read_csv("_data/formula1/drivers.csv")
drivers <- drivers %>%
  mutate(number = na_if(number, "\\N"))

# Lap Times
lap_times <- readr::read_csv("_data/formula1/lap_times.csv")
lap_times <- lap_times %>%
  mutate(laptime_seconds = milliseconds/1000) %>%
  select(-c(time, milliseconds))

# Pit Stops
pit_stops <- readr::read_csv("_data/formula1/pit_stops.csv")

# Qualifying
qualifying <- readr::read_csv("_data/formula1/qualifying.csv")
qualifying <- qualifying %>%
  select(-number) %>%
  mutate(across(q1:q3, ~ period_to_seconds(ms(na_if(.x, "\\N"))))) %>%
  rename(q1_time_s = q1, q2_time_s = q2, q3_time_s = q3)

# Races
races <- readr::read_csv("_data/formula1/races.csv")
races <- races %>%
  select(raceId:name) %>%
  rename(gp_name = name)

# Results
results <- readr::read_csv("_data/formula1/results.csv")
results <- results %>%
  mutate(
    disqualified = as_factor(ifelse(positionText == "D", 1, 0)),
    retired = as_factor(ifelse(positionText == "R", 1, 0)),
    fastestLap = as.numeric(fastestLap),
    rank = as.numeric(rank),
    fastestLapTime = period_to_seconds(ms(na_if(fastestLapTime, "\\N"))),
    fastestLapSpeed = as.numeric(fastestLapSpeed),
    finishTimeSeconds = as.numeric(na_if(milliseconds, "\\N"))/1000) %>%
  rename(start_position = grid, finish_position = positionOrder) %>%
  select(-c(number, position, positionText, time, milliseconds)) 

# Seasons
seasons <- readr::read_csv("_data/formula1/seasons.csv")

# Status
status <- readr::read_csv("_data/formula1/status.csv")

```


### Dealing With Changing Team Names

An important note that was not considered in the previous homework is that in order to look at details by team, we will need to take another consideration into account: team names. From 2016 - 2022, there are quite a few different team names recorded in the data:

```{r}
constructor_standings %>%
  inner_join(constructors, by = 'constructorId') %>%
  filter(raceId >= 948) %>%
  select(constructor_name) %>%
  unique()
```


Although these teams all have separate names and designations, some of them are actually the *same* team, just in different years. For example, the current "Alpha Tauri" team used to be called "Toro Rosso". In order to keep the teams joined together through name changes, we can recategorize the team names as just their current name. 

Additionally, we are interested in comparing things before and after the rules change in 2022. This is impossible for teams that no longer existed in any form in 2022. The only team fitting in this category is "Manor Marussia". We can simply remove this one from our data. 


```{r}
constructors_renamed <- constructors %>%
  filter(constructor_name %in% c("AlphaTauri", "Toro Rosso", "Alfa Romeo", "Sauber", "Alpine F1 Team", "Renault", "Aston Martin", "Racing Point", "Force India", "Haas F1 Team", "Ferrari", "McLaren", "Mercedes", "Red Bull", "Williams")) %>%
  mutate(constructor_name = 
           case_when(
             constructor_name %in% c("AlphaTauri", "Toro Rosso") ~ "Alpha Tauri",
             constructor_name %in% c("Alfa Romeo", "Sauber") ~ "Alfa Romeo",
             constructor_name %in% c("Alpine F1 Team", "Renault") ~ "Alpine",
             constructor_name %in% c("Aston Martin", "Racing Point", "Force India") ~ "Aston Martin",
             constructor_name == "Haas F1 Team" ~ "Haas",
             TRUE ~ constructor_name #everyone else has remained the same name
             
         )) 
```



## Research Questions

Using this data, I'd like to investigate how Formula 1 has shifted before and after the most recent regulation change in 2022.

This change shifted how F1 cars look and perform, with new engine requirements, weight requirements, and aerodynamic guidelines. Just by watching (and not looking at any data) this change seems to have shaken things up considerably, finally ending the Mercedes' team's dominant win streak in the Constructors championship and seeing Red Bull take first place instead.

I'm interested in using this dataset to take a deeper look into how things have changed since the regulations shift -- comparing the previous era to the new era, which started in 2022. 

There are a few different aspects of this this comparison that are of interest:

1. The new regulations were designed to make it easier for cars to follow one another, theoretically making passing easier. How has this translated in reality? Is it easier now than before for "top" cars that start out of position to make up positions? Being "out of position" here means starting the race from a position worse than would be expected based on their car's performance due to a qualifying mistake, mechanical failure, etc.

2. How have lap times changed, both in qualifying and the race? Are the new cars getting faster/slower overall? Is this change the same across all race tracks?

3. How has car reliability changed? Are mechanical failures in races more common or less common than before? Are certain types of failures popping up that we didn't see much of before (or the opposite)?

4. Which drivers did the best job adapting to the regulations? We could look into how lap times changed overall on average, and then compare how close each driver's lap time changes were to the mean. 

5. How did the regulations shake up the order in the Constructors championships? Did any teams go from dominant to mediocre? Did anyone suddenly shoot up?

We will look at these questions one by one.

Note: The previous era prior to the regulation change, called the Turbo Hybrid era, technically started in 2014. However, due to a shift in qualifying formats in 2016, we will ignore 2014/15 and just look at 2016-2021 ("Before") and 2022 ("After") as our two comparison periods. We will also exclude the 2023 season as it is still not complete and data is not properly updated yet.


### **Is it easier to pass?**

For this data, we need to look at the `results` dataset and capture start position, finish position, and overall positions changed for the time period in question. 

```{r}
positions_gained <- results %>%
  filter(raceId >= 948) %>%
  select(raceId, driverId, constructorId, statusId, start_position, finish_position) %>%
  mutate(positions_gained = start_position - finish_position)
```

We then join this information to the race info, as well as driver and constructor names:

```{r}
positions_clean <- positions_gained %>%
  inner_join(drivers %>% select(driverId, code, surname), by = 'driverId') %>%
  inner_join(constructors_renamed %>% select(constructorId, constructor_name), by = 'constructorId') %>%
  inner_join(races %>% select(raceId, year, gp_name), by = 'raceId') %>%
  select(code, surname, constructor_name, year, gp_name, start_position, finish_position, positions_gained, statusId)
  
positions_clean %>% head(5)
```

Taking a quick look at our new positions gained variable, we can get some basic descriptive statistics:

```{r}
summary(positions_clean$positions_gained)
```

The max number of positions lost is 22 (not sure how that is possible as there are only 20 spots on the grid?). Max gained is 18, which is very impressive. On average, people lose 0.3 places.

In order to see if it is easier to pass, we can look into how much the average # of positions gained per race has changed over time (specifically for those who finished the race):

```{r}
positions_clean %>%
  filter(statusId == 1) %>%
  group_by(year) %>%
  summarize(avg_gain = mean(positions_gained)) %>%
  mutate(post_rules_change = ifelse(year == 2022, TRUE, FALSE)) %>%
  ggplot(aes(x = year, y = avg_gain, fill = post_rules_change)) + 
  geom_col() +
  ylab("Average Positions Gained") +
  xlab("Season") + 
  ggtitle("Avg Positions Gained For Drivers Who Finished, 2016-2022") + 
  scale_fill_manual(values = c("gray", "orange")) +
  theme_minimal()
```
Overall, we see a jump from 2021 (before the rules change) to 2022 (after), but it is very small and doesn't seem to be part of a larger trend.

What if we just look at cars that started out of position? For this analysis, we will define this as a car that **started at least 2 places behind the average starting position of their team that year**. 

We can calculate average start position for each team each year and use that to identify who is starting out of position:

```{r}
avg_start <- positions_clean %>%
  group_by(year, constructor_name) %>%
  summarize(avg_start = mean(start_position))

out_of_position <- positions_clean %>%
  inner_join(avg_start, by = c('year', 'constructor_name')) %>%
  mutate(out_of_position = ifelse((start_position - avg_start >= 2), TRUE, FALSE))

out_of_position %>%
  group_by(year, out_of_position) %>%
  summarise(avg_gain = mean(positions_gained)) %>%
  mutate(post_rules_change = ifelse(year == 2022, TRUE, FALSE)) %>%
  ggplot(aes(x = year, y = avg_gain, fill = out_of_position, color = post_rules_change)) + 
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0) + 
  ylab("Average Positions Gained") +
  xlab("Season") + 
  ggtitle("Avg Positions Gained Based on Out Of Position Start, 2016-2022") + 
  scale_fill_manual(values = c("gray", "lightblue")) +
  scale_color_manual(values = c("white", "orange")) +
  theme_minimal()
```
We see now that drivers who started in position on average *lost* positions (but typically fewer) while drivers who started out of position on average *gained* positions (typically more). This makes sense logically. When a fast car starts the race behind where they should, they can more easily pass. If a car starts at its average position or even higher, then it is surrounded by other cars that are at least as good, and so it's much easier to be passed and much harder to pass.

Looking at the trend overall, there does not seem to be a clear change with the 2022 rule change. In 2022, the average number of positions gained for cars starting out of position was higher than 2021, but is more so part of a continuing upward trend than an outlier, particularly as we note that both 2017 and 2018 have higher average positions gained.

For cars starting in position, we see a similar trend of changing times. However, we do note that 2022 is the largest number of positions lost on average across all years in our time frame, so it is possible that starting in position (or ahead of expectations) left drivers more vulnerable than in past years. In other words, just because they started at the front does not mean that a faster car behind could not pass.

> == COMMENTARY == : 
> I think that the second graph definitely helps provide context that the first one is missing (by adding in whether a driver started out of position). It might be slightly confusing though that in a sense, a driver could also be "out of position" by starting AHEAD of where they would be expected, so at the moment we are more so comparing starting behind expectation to starting at or above expectation. Not sure if this makes sense, or if it would make more sense to categorize as below expecation / at expectation / above expectation.



### **How have lap times changed?**

Next, we'll look at lap times to see if there were any noticeable changes. To look at how lap times changed before and after the rules change, we need to consider the shifting calendar. The races that happen each year are not always the same. For example, 2021 had a Grand Prix in Russia, but this race was cancelled in 2022 due to the war in Ukraine and a ban on Russian participation in motorsport racing.

In order to look at changes over time, we will just look at the races that happened in 2022 and look at how those specific tracks changed, ignoring races in the previous years on tracks that were not used in 2022.

Since there are so many tracks, we'll just select a random set of 5 to look at, and we'll also remove outlier laps over 150 seconds.

```{r}
tracks_2022 <- races %>%
  filter(year == 2022) %>%
  select(gp_name)

selected_laps <- lap_times %>%
  inner_join(drivers %>% select(driverId, code, surname), by = 'driverId') %>%
  inner_join(races %>% select(raceId, year, gp_name), by = 'raceId') %>%
  filter(raceId >= 948, gp_name %in% c(tracks_2022$gp_name)) %>%
  mutate(post_rules_change = ifelse(year == 2022, TRUE, FALSE))
```

```{r}
set.seed(4)
races_list <- sample(tracks_2022$gp_name,5)

selected_laps %>%
  filter(gp_name %in% races_list, laptime_seconds < 150) %>%
  ggplot(aes(x = laptime_seconds, fill = post_rules_change)) + geom_density(alpha = 0.5) + 
  facet_wrap(~gp_name) +
  scale_fill_manual(values = c("white", "orange")) + theme_minimal() +
  xlab("Lap Time (s)") + ggtitle("Lap Time Distributions For Select GPs, Pre and Post 2022 Rule Change")
```
There is no clear pattern here.

For the Australian GP, lap times appear to be significantly slower after the rules change than prior. The French GP seems to be slightly faster, as does the Austrian GP. The US Grand Prix times seem pretty much the same. So it is possible that the regulations may have made lap times change differently on different tracks.  

We can conduct a quick T-test to see if, controlling for track, we see a difference in track times before and after the rules change:

```{r}
model <- lm(laptime_seconds ~ post_rules_change + gp_name, data = selected_laps)
summary(model)
```

Here, controlling for the race name, we see the following information for the rules change variable:

> post_rules_changeTRUE  ||  Estimate: 1.2066  ||  P-value: 0.0437 *  

The p-value of < 0.05 indicates the presence of a potential relationship between lap times and the rules change. Since the estimate is positive, this means that statistically, the lap times seem to be *slightly longer* after the rules change than before, controlling for the track.

We can also do some further analysis to determine which tracks had the greatest change in lap time by looking at their average lap time before the change and comparing to 2022 times. We display the 5 tracks with the greatest **increase** in lap times, followed by the 5 with the greatest **decrease**, again removing outlier laps over 150 seconds which would indicate some sort of car or weather related issue:

```{r}
laptime_changes <- selected_laps %>%
  filter(laptime_seconds < 150) %>%
  group_by(gp_name, post_rules_change) %>%
  summarize(avg_lap = mean(laptime_seconds)) %>%
  mutate(post_rules_change = ifelse(post_rules_change == TRUE, "after_change", "before_change")) %>%
  pivot_wider(names_from = post_rules_change, values_from = avg_lap) %>%
  mutate(difference = after_change-before_change, direction = ifelse(difference < 0, "Faster", "Slower"))


laptime_changes %>%
  filter(direction == "Faster") %>%
  arrange(difference) %>%
  head(5)

laptime_changes %>%
  filter(direction == "Slower") %>%
  arrange(desc(difference)) %>%
  head(5)
```

We can also plot this on ggplot to get a sense overall:

```{r}
laptime_changes %>%
  filter(!is.na(difference)) %>%
  ggplot(aes(x= reorder(gp_name, difference), y = difference, fill = direction)) + geom_col() + coord_flip() + scale_fill_manual(values = c( 'lightgreen', 'tomato')) +
  xlab("GP Name") + ylab("Avg Lap Time Difference") + ggtitle("Difference in Avg Lap Time, Pre- and Post- 2022 Rules Change") +
  theme_minimal()
```
Here we can see that more tracks had *slower* lap times than had *faster*. We can also see that in the "Faster" direction, the Abu Dhabi GP is an outlier, and most tracks only got *slightly* faster if they did at all. In the "Slower" drection, there were 3 tracks (Japan, Singapore, Monaco) that were > 5 seconds slower. 

> == COMMENTARY == : 
> For our first graph with the 5 random density plots, it is hard to get a sense of the actual pattern so I'm not sure really how helpful this visualization is. I tried doing a version with all the tracks but it was just far too chaotic. Not sure if it is worth keeping some version of this, but potentially having more reasoning behind which tracks were selected, versus just a random draw? But not sure how to define the selection criteria.
> The statistical model definitely helps make this trend more clear though!
> I really like the graph with the red and green lap time changes. One thing that could maybe be added to make this more interesting though would be to classify each track further by track type in some way. For example, street track vs purpose built track. These tend to have very different characteristics that could provide further insight
> It might also make sense to look at this data by constructor to see which constructors had the biggest jump in lap times


### **How has reliability changed?**

### **Which drivers adapted best to the regulations?**

### **How did the regulation change shake up the championship?**

For this question, we will look at the Constructors Standings.

Do we see a significant change in the lineup of teams before and after the change? To determine this, we can create a timeseries plot showing how team placings changed season by season from 2016-2022.

We start by joining the constructors and races tables together:

```{r}
# Join constructor results table to more information on constructors and races
constructor_results <- constructor_standings %>% 
  inner_join(constructors_renamed, by = 'constructorId') %>%
  inner_join(races, by = 'raceId')

head(constructor_results)
```

We can then filter this to the years and columns relevant to us, and group by/summarize to find the total number of points each team got each season:

```{r, message = FALSE}
constructors_ranked <- constructor_results %>%
  filter(year >= 2016 & year < 2023)  %>%
  select(year, raceId, position, points, constructor_name) %>%
  group_by(year, constructor_name) %>%
  summarize(final_points = max(points))
```

Using the same historic team name update as we have done previously, we will use reconfigured team names (which use the most recent name to reference all iterations of the team) to make it easier to compare each team's performance over time. We also include a vertical line to show the point where the transition occurred. The line and to the left are before the transition. To the right is after.

Although these teams all have separate names and designations, some of them are actually the *same* team, just in different years. For example, the current "Alpha Tauri" team used to be called "Toro Rosso". In order to keep the teams joined together through name changes, we can recategorize the team names as just their current name and also remove the teams that no longer existed in the 2022 season. This helps make our graph easier to follow so we can better trace the changes over time.


```{r}
constructors_ranked %>%
  ggplot(aes(x = year, y = final_points, color = constructor_name)) + 
  geom_line() + 
  geom_vline(xintercept = 2021) + 
  theme_minimal() +
  xlab("Season") +
  ylab("Constructors' Championship Points") +
  ggtitle("Points Earned In The Constructors' Championship, 2016-2022", subtitle = "Using Current Team Names")
```

As we can see here, relative rankings in the championship remained pretty steady for a good part of the old regulations, with Mercedes dominant, Ferrari having a few good years, and Red Bull close behind. 

After the regulations change, Mercedes dropped from 1st to 3rd place and Red Bull and Ferrari had massive spikes in points. However, note that they were on the up-swing before the change was put in place (in 2021) so it could be unrelated. McLaren did not seem to take the change well. They were previously on an upward points scoring trend and shot down in 2022. We see the same trend with Alpha Tauri. Meanwhile, Hass and Alfa Romeo, who in previous years had struggled to score any points did see a small jump in points and a positive shift in their overall rankings.

It is hard to tell fully, without a few more years worth of data, but it does appear that the regulation change did switch the team rankings up and some teams (Red Bull/Ferrari) adapted better than others (Alpha Tauri, McLaren).

We can also look at the number of wins and podiums by each team, where here we define podiums as a 1st, 2nd or 3rd place finish (which includes the 1st place finishes from the wins column).

```{r message=FALSE}
wins_podiums <- constructor_standings %>% 
  inner_join(constructors_renamed, by = 'constructorId') %>%
  inner_join(races, by = 'raceId') %>%
  inner_join(results, by = c('raceId', 'constructorId')) %>%
  select(constructor_name, year, gp_name, rank) %>%
  filter(year >= 2016 & year < 2023) %>%
  group_by(constructor_name, year) %>%
  summarize(wins = sum(rank == 1), podiums = sum(rank %in% c(1, 2, 3)))

wins_podiums %>%
  arrange(desc(podiums))
```

Looking at a summary of each of these new variables we see that for wins, the most we had recorded was 10 in a season. The average was 2 wins per team.

```{r}
summary(wins_podiums$wins)
```

For podiums, the max was 29 across both drivers, with an average of 6 per team.

```{r}
summary(wins_podiums$podiums)
```

```{r}
wins_podiums %>%
  ggplot(aes(x = year, y = wins, col = constructor_name)) + geom_line() + theme_minimal() + ggtitle("Wins Per Season, 2016-2022") + geom_vline(xintercept = 2021) 
```

We see that looking at Wins, Mercedes took a big dip and Ferrari took a big jump, while Red Bull remained pretty steady after the rule change.

```{r}
wins_podiums %>%
  ggplot(aes(x = year, y = podiums, col = constructor_name)) + geom_line() + theme_minimal() + ggtitle("Podiums Per Season, 2016-2022", subtitle = "Where a Podium Here Only Includes 2nd/3rd") + 
  geom_vline(xintercept = 2021) 
```

Looking at podiums, we see a similar trend.

> == COMMENTARY == : 
> It is kind of difficult to tell all of these team colors apart. Would it make sense to use an official color palette that is representative of the actual team names? Might be worth adding for a final version.
> Not sure if it would be interesting or not to look at average finish position as well versus just points (from our first graph). If we look at points, we are missing all the information happening between 11th-20th place since no points are awarded. Technically a team could be finishing middle of the road in 11th all season and they'd look just as bad on the points time table as a team finishing 20th every time, so we definitely lose some info if we don't look at finish position. 

## References


#### Datasets

Rao, R. (2023). Formula 1 World Championship Data: 1950-2023. *Kaggle*.
<https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020>

Formula 1 (2023). Formula 1 Race Result Archive. *Formula 1*.
<https://www.formula1.com/en/results.html>


#### Overview of F1

Federation International de l'Automobile (FIA). (2023). Regulations: FIA Formula One World Championship. 
<https://www.fia.com/regulation/category/110>

Tippett, B. (2021). The Complete Guide To Understanding Formula 1. *Defector*.
<https://defector.com/the-complete-guide-to-understanding-formula-1>

Kanal, S. (2023). The beginner's guide to...the Formula 1 Grand Prix Weekend. *Formula 1*.
<https://www.formula1.com/en/latest/article.the-beginners-guide-to-the-formula-1-grand-prix-weekend.20OGbgZCWKj9ML79gBzfoX.html>

F1 Chronicle Media Team (2020). The Complete Beginners Guide to Formula 1. *F1 Chronicle*.
<https://f1chronicle.com/a-beginners-guide-to-formula-1/>

#### Techincal / Software

Wickham, H., & Grolemund, G. (2016). R for data science: Visualize, model, transform, tidy, and import data. 
<https://r4ds.had.co.nz> 

R Core Team (2023). _R: A Language and Environment for Statistical Computing_. R Foundation for Statistical Computing, Vienna,
  Austria. <https://www.R-project.org/>.




