---
title: "Homework 2"
author: "Audrey Bertin"
description: "Reading in final project data - Formula 1"
date: "6/9/2023"
format:
  html:
    df-print: paged
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - hw2
  - audrey_bertin
  - formula_1
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


For my final project, I'm interested in working with data from Formula 1 (the motorsport racing series).

## What is F1?

Formula 1 (F1) is the top class of single-seater motorsport racing. "Formula" refers to the set of rules to which all participants' cars must conform. For example, Formula 2 has different rules about car design and racing than Formula 1.

An F1 season consists of a series of races, known as Grands Prix (GP), which take place around the world on purpose-built racing circuits as well as public roads (termed "street circuits"). The number and location of races per season has varied significantly across the years.

There are currently 10 teams in Formula 1, officially termed "Constructors". Like the races, the number of teams (and names) has varied significantly over time. Each team has *two* drivers who race for them each season (as well as a driver on standby in case one of the primary drivers get sick/injured). 

Each Constructor is responsible for designing their own car each season within a shared set of regulations that restrict the design across all teams (e.g. a minimum weight, certain aerodynamic requirements, etc.). Some Constructors are known as "works teams". These teams design and construct the engines for their cars *in addition* to the aerodynamic components. The rest of the teams are called "customer teams". They design the aerodynamic parts of their car, but *purchase* their engines from one of the works teams instead of designing it themselves. For example, currently, the Mercedes *works* team uses their engines for their own car and  *also* provides engines to the Williams and McLaren teams. 

There is also the concept of "junior teams", where one company essentially has ownership over multiple teams in F1. They have one primary, A-team, and then a B-team that they can use to develop talent for their A-team. For example, currently Red Bull owns both the Red Bull Racing *and* Alpha Tauri teams, and regularly promote their drivers from Alpha Tauri to Red Bull Racing. For the purposes of the championship, however, these are considered *separate* constructors.

Because each Constructor designs their own car, cars can vary quite dramatically across teams, so there tends to be a somewhat clear hierarchy. The "worst" car often has nearly zero chance of winning a race, and the "best" cars almost always end up on the podium. In formula 1 lingo, teams are broken into three categories:

- "top" teams -- these teams win most of the races / usually make up the podium
- "midfield" teams -- these teams are not as fast as the top teams but often still place reasonably well and can, on occasion, end up on the podium or even (in rare cases) win a race
- "backmarker" teams -- these teams are significantly slower than the others, with *almost no* chance of ending up on the podium, and usually finish near the back of the pack

Due to this team hierarchy, it can be difficult to compare a driver from one team to a driver on another. Just because one driver always has faster lap times or finishes in a better position does not mean that they are necessarily a better driver, and could just mean that they have a better car. Since both teammates on a given team drive the same car, driver skill is often judged more-so based on how well a driver does compared to their teammates. 

In each season, there are two separate championships up for grabs -- a **Drivers Championship** and a **Constructors Championship**. The Drivers Championship is considered more prestigious and awards the individual driver who received the most points throughout the season. The Constructors Championship is awarded to the *team* with the most *combined* points across their two drivers. 

Teams and drivers have the chance to earn points at each Grand Prix (note that there are also points available in what are called "Sprint races", but these are a new feature in F1 and we'll be excluding them for the purposes of this analysis in order to simplify things).

Each GP occurs over one weekend is broken up into two segments: qualifying (typically on Saturday) and the race (typically on Sunday). In qualifying, drivers try to put together the *fastest* lap. Lap times then determine the order that the cars will start for the race.

Within the race, cars are required to make *at least one* pit stop to change which type of tires they are using (there are hard, medium, and soft tires available for each race, each of which have different pros and cons). Depending on how much a given track damages the tires, along with other factors such as if a car is involved in a collision, additional pitstops may take place to change tires or swap out parts.

At the end of the race, points are awarded based on drivers' finishing position. Additionally, the driver who records the fastest single lap throughout the race is awarded an extra point in addition to their points for finishing position. 

The current points distribution is as follows:

- 1st place: 25 pts
- 2nd place: 18 pts
- 3rd place: 15 pts
- 4th place: 12 pts
- 5th place: 10 pts
- 6th place: 8 pts
- 7th place: 6 pts
- 8th place: 4 pts
- 9th place: 2 pts
- 10th place: 1 pt
- Fastest Lap: 1 pt 
- 11th-20th place: 0 pts

Points are cumulative across a season, and the championships are awarded to the team and driver with the highest point totals.

## Dataset

The dataset I'll be using is hosted on Kaggle at the following link: https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020

It contains F1 race data from the series' start in 1950 to the present season, which is currently in-progress and will run until November.

There are 14 separate files within this dataset:

- `circuits.csv` contains information about the different tracks the drivers race on (a row here is a track)
- `constructor_standings.csv` contains information about how many points each team has after each race (a row here is a single team at a single race)
- `constructors.csv` stores the names and nationalities of teams (a row here is a team)
- `driver_standings.csv` stores information about how many points each driver was awarded at each race (a row here is a single driver at a single race)
- `drivers.csv` stores the names and nationalities of drivers (a row here is a driver)
- `lap_times.csv` stores the lap times for every driver on every lap of each race (a row here is a single lap for a single driver at a single race)
- `pit_stops.csv` stores information about every pit stop taken during each race (a row here is a single pit stop taken by a single driver at a single race)
- `qualifying.csv` stores qualifying lap times for each driver from each GP (a row here is a single driver in a single GP, and each row contains all qualifying lap times across multiple sessions)
- `races.csv` contains information about each GP, e.g. track name, date, etc. (a row here is a single GP)
- `results.csv` contains all of the finishing results for each race, e.g. driver finishing positions, # laps completed, fastest lap (a row here is a single driver at a single race)
- `seasons.csv` contains the years and urls for wikipedia pages associated with each season (a row here is a season)
- `sprint_results.csv` same as results.csv but for sprint races, which we'll be excluding (a row here is a single sprint race)
- `status.csv` contains the encodings of a `status` variable that is referenced in the results files (a row here is a status)


## Data Cleaning

We now look at each of these one by one, excluding the sprint race results.

#### `circuits.csv`

```{r message=FALSE}
circuits <- readr::read_csv("_data/formula1/circuits.csv")
```

```{r}
head(circuits)
```

Each circuit has a `circuitId` that uniquely identifies it. We also have location information. The column `alt` is not entirely clear. 

All of the data types seem reasonable here so there is no need to modify them. However, we have a bunch of data frames that are all going to have a column called `name` as we'll see later. We should rename these columns to avoid confusion when we join tables in the future.

```{r}
circuits <- circuits %>%
  rename(circuit_name = name)

colnames(circuits)
```





#### `constructor_standings.csv`

```{r message=FALSE}
constructor_standings <- readr::read_csv("_data/formula1/constructor_standings.csv")
```

```{r}
head(constructor_standings)
```

The constructors standings data has a unique identifier for each row, as well as an id that refers to the races df as well as the constructors df. For each constructor/race we see how many points they have after that race, their current standing in the championship, as well as their running tally of wins in the season so far. 

All of the data types seem reasonable here as well. `constructorId` would make more sense as a factor, but we would want to wait until later on when we join tables together to make this update since this refers to a column in another table. For now, `positionText` seems to duplicate `position` so we can remove `positionText`.

```{r}
constructor_standings <- constructor_standings %>%
  select(-positionText)

constructor_standings
```



#### `constructors.csv` 

```{r message=FALSE}
constructors <- readr::read_csv("_data/formula1/constructors.csv")
```

```{r}
constructors
```
For each constructor we have a unique `constructorId` (211 of them), which can be used to join this data with other tables. We also get names, nationalities and wikipedia URLs.

All of the data types are reasonable here. Converting team names to factors would again be better after completing joins.

We can again rename the `name` column to avoid confusion later.

```{r}
constructors <- constructors %>%
  rename(constructor_name = name)

colnames(constructors)
```


#### `driver_standings.csv` 

```{r message=FALSE}
driver_standings <- readr::read_csv("_data/formula1/driver_standings.csv")
```

```{r}
driver_standings
```
This file has the exact same format as the constructors standings. A unique ID for each row, a race ID and driver ID to join with other tables, cumulative points, race position, and a running count of wins.

Again, `positionText` duplicates `position` so we can remove it.

```{r}
driver_standings <- driver_standings %>%
  select(- positionText)

driver_standings
```



#### `drivers.csv` 

```{r message=FALSE}
drivers <- readr::read_csv("_data/formula1/drivers.csv")
```

```{r}
drivers
```

We have 857 unique drivers. Each driver has a unique ID, a race number, a 3 letter code used to refer to them on score sheets/race documentation, a full name, DOB, nationality, and wikipedia link.

Driver number seems like it should be numeric, but it makes sense to leave as a character because the numbers aren't actually ordered in any way (drivers choose them based on personal preference). There does seem to be a strange missing data marker (`\\N`), so we will replace this with a normal `NA`.

```{r}
drivers <- drivers %>%
  mutate(number = na_if(number, "\\N"))

drivers
```


#### `lap_times.csv` 

```{r message=FALSE}
lap_times <- readr::read_csv("_data/formula1/lap_times.csv")
```

```{r}
lap_times
```

Lap times has unique identifiers referring to the race and driver who completed them, which we can use to join tables later. We also have lap number, driver position, and lap times. The `time` variable is in an `hms` class format, but to make analysis easier, we'll get rid of it and just use the `milliseconds` column converted to seconds.

```{r}
lap_times <- lap_times %>%
  mutate(laptime_seconds = milliseconds/1000) %>%
  select(-c(time, milliseconds))

lap_times
```


#### `pit_stops.csv` 

```{r message=FALSE}
pit_stops <- readr::read_csv("_data/formula1/pit_stops.csv")
```

```{r}
pit_stops
```

For the pit stops, we have a unique race and driver ID to join with other tables. We also get the stop number for that driver within the race, the lap in which the stop was taken, time of day and pit stop duration.

The only column with obviously the wrong data type is duration. Since this measures time, it should be a double. We can also rename duration to `duration_seconds` to make the units more obvious. 

```{r}
pit_stops <- pit_stops %>%
  mutate(duration = as.numeric(duration)) %>%
  rename(duration_seconds = duration)

pit_stops
```

#### `qualifying.csv` 

```{r message=FALSE}
qualifying <- readr::read_csv("_data/formula1/qualifying.csv")
```

```{r}
qualifying
```

For qualifying, we get a unique race, driver, and constructor ID. We also have a column called "number" which is not entirely clear what it refers to. We then have overall qualifying position and qualifying times in the three different sections of qualifying (q1/q2/q3).

Since number is not clear, we can remove it. `q1/q2/q3` are all times and should be converted as such. To make this easy, we'll just convert the times to seconds. We'll then rename the columns slightly to make them more clear. In the process, we'll also replace the weird NA characters with normal NA values.

```{r}
qualifying <- qualifying %>%
  select(-number) %>%
  mutate(across(q1:q3, ~ period_to_seconds(ms(na_if(.x, "\\N"))))) %>%
  rename(q1_time_s = q1, q2_time_s = q2, q3_time_s = q3)

qualifying

```

#### `races.csv` 

```{r message=FALSE}
races <- readr::read_csv("_data/formula1/races.csv")
```

```{r}
races
```

This table contains information about each race -- year, round (race number within year), circuit, race name, date, time, url, and some blank columns. For the purpose of analyzing anything about the sport, race dates and times are not relevant, so we'll remove them. We can also remove all of the blank columns.

Again, we see the `name` variable, so we'll rename this as well.

```{r}
races <- races %>%
  select(raceId:name) %>%
  rename(gp_name = name)

races
```

#### `results.csv` 

```{r message=FALSE}
results <- readr::read_csv("_data/formula1/results.csv")
```

```{r}
results
```

For race results, we see a race, driver, and constructor ID we can use to join other tables with. Again, we see a column called number that we don't know what it means, so we can remove it.

`grid` represents starting position, so we can rename it as such (`start_position`).

For position, we have three different columns that are telling us something similar (almost duplicates). `positionText` is helpful because it shows us information such as whether the driver was disqualified ("D") or whether they retired from the race ("R"). `positionOrder` is also helpful as this is the official classification used by F1 (D/R don't matter for the classification). Rather than keep both `positionText` and `positionOrder`, we'll keep `positionOrder` and then create new indicator columns for retirements and disqualifications. We will also rename `position_order` to `finish_position` to be more accurate.

The character representations of time are useless for analysis, so we'll just keep milliseconds but convert to seconds. We will also make this a numeric column instead of character.

Fastest lap and rank should also be numeric, as should fastest lap speed. Fastest lap time should be converted to seconds like we did before.

Status ID is fine as is.

We make all of these modifications below:

```{r}
results <- results %>%
  mutate(
    disqualified = as_factor(ifelse(positionText == "D", 1, 0)),
    retired = as_factor(ifelse(positionText == "R", 1, 0)),
    fastestLap = as.numeric(fastestLap),
    rank = as.numeric(rank),
    fastestLapTime = period_to_seconds(ms(na_if(fastestLapTime, "\\N"))),
    fastestLapSpeed = as.numeric(fastestLapSpeed),
    finishTimeSeconds = as.numeric(na_if(milliseconds, "\\N"))/1000) %>%
  rename(start_position = grid, finish_position = positionOrder) %>%
  select(-c(number, position, positionText, time, milliseconds)) 

results
```


#### `seasons.csv` 

```{r message=FALSE}
seasons <- readr::read_csv("_data/formula1/seasons.csv")
```

```{r}
seasons
```

The seasons data is very simple and just has a year and wikipedia URL. There is no need to modify anything here, and this data overall does not look very useful.

#### `status.csv` 

```{r message=FALSE}
status <- readr::read_csv("_data/formula1/status.csv")
```

```{r}
status
```


In the `results` data frame there is a column called status ID:

```{r}
results %>%
  select(resultId, raceId, driverId, constructorId, start_position, finish_position, statusId)
```


The `status` table contains the encodings for this table. For example, the first 5 status IDs are all 1, which indcates "Finished", while the two people with status 5 had "Engine" issues.

We can quickly see what the top statuses are:

```{r message = FALSE}
top_statuses <- results %>%
  inner_join(status,by="statusId") %>%
  group_by(statusId, status) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) 
```

```{r}
top_statuses %>%
  head(15) %>%
  ggplot(aes(x=reorder(status,n), y=n)) + geom_col() +
  ggtitle("Most Common Race Statuses by Count") + theme_minimal() + xlab("Status") + ylab("Count") + coord_flip()
```



Most drivers finished. A good number were lapped. Most common reasons for DNF appear to be engine failure, accident, and collision.

As we can see, the status table will only be useful if joined with the results table in future analysis.

We don't need to change anything about the data types in the status table. They all look good.

## Research Questions

For my final project, I'm specifically insterested in the driver *Max Verstappen* (Driver ID: 830, Constructor IDs: 5 -- Toro Rosso & 9 -- Red Bull), who has won the last two Drivers Championships, in 2021 and 2022. From purely an observational (and non-data-based) standpoint, he seems like he is in a league above everyone else in F1 right now, and I'm curious to see how the data stacks up and where his strengths are. Some specific areas to look into are:

- **Pole to Win Ratio:** How often does Max convert pole position (starting first) to winning a race, compared with other drivers with lots of pole positions?
- **Lap Times:** How does Max's lap time (both in qualifying and in the race) compare to his teammates?
- **Wins and Podiums:** Max seems to be picking up wins and podiums at an *insane* rate. How does this rate compare to other great drivers like Michael Schumacher or Lewis Hamilton?
- **Points:** The constructors championship is based on the combined points earned by both drivers. What proportion of these points is Max bringing in for his team, and how does this stack up to other teammate pairs?
- **Rise to Success:** How has Max's performance changed since joining the sport?
- **Come-From-Behinds:** Max seems to have an insane ability to win, even if he's starting a race at the back of the grid (i.e. in 11th-20th place). How does this ability stack up to other drivers?
- **DNFs:** How often does Max DNF a race? And how much of the time is it *his* fault, versus a mechanical failure?

For the purposes of this analysis, I will likely just look at drivers who raced at least a season between 2006 and 2022, including those whose first season was prior to 2006. Why start at 2006? Because this was a huge regulations change year which shifted the car design dramatically, toward what we have today, and comparing across historical eras that are *too* different can make numbers pretty meaningless.

This analysis will likely be pretty complicated, involving a lot of table joining, but the data looks to be in a good format for this sort of joining with all the unique identifier columns in each table.



