---
title: "Challenge 9"
author: "Audrey Bertin"
description: "Creating a function"
date: "5/30/2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_9
  - functions
  - audrey_bertin
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

For this week's challenge, I'll create a function that calculates z score for a certain value based on a baseline computed from a vector.

It will take the following as input:

- `baseline` = vector of numbers of any length
- `value` = the value to compute the z score for

It will then give the following as output:

- a dataframe with `mean`, `sd` (of baseline), `input_value` and `z_score`

#### Function Definition


```{r}
z_score <- function(baseline, value){
  mean <- mean(baseline)
  sd <- sd(baseline)
  z_score <- abs((value - mean)/sd)
  
  results = tibble(mean = mean, sd = sd, input_value = value, z_score = z_score)
  return(results)
}
```

#### Testing


Let's use the following inputs to test the function:

```{r}
baseline = c(4, 2, -1, 4, 9, 2, 3, 3, 1 -5)
value = 10
```


We expect the following results if this runs correctly:

- Mean should be: `r mean(baseline)`
- SD should be: `r sd(baseline)`
- Input value should be `r value`
- Z score should be the absolute value of (`value` - `mean`) / `sd` -> abs value of (`r value` - `r mean(baseline)`) / `r sd(baseline)` -> `r abs((value -  mean(baseline)) / sd(baseline))`

We confirm this by running the function below and confirming the results match:

```{r}
results = z_score(baseline, value)
results
```

```{r}
expected_mean = mean(baseline)
actual_mean = results$mean

expected_sd = sd(baseline)
actual_sd = results$sd

expected_input = value
actual_input = results$input_value

expected_z = abs((value -  mean(baseline)) / sd(baseline))
actual_z = results$z_score

paste0("Mean matches: ", expected_mean == actual_mean)
paste0("SD matches: ", expected_sd == actual_sd)
paste0("Input matches: ", expected_input == actual_input)
paste0("Z Score matches: ", expected_z == actual_z)
```







