---
title: "Challenge 5 Solution"
author: "Shreya Varma"
description: "Introduction to Visualization"
date: "5/30/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_5
  - cereal
  
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least two univariate visualizations
   - try to make them "publication" ready
   - Explain why you choose the specific graph type
5)  Create at least one bivariate visualization
   - try to make them "publication" ready
   - Explain why you choose the specific graph type

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code.

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   cereal.csv ⭐
-   Total_cost_for_top_15_pathogens_2018.xlsx ⭐
-   Australian Marriage ⭐⭐ 
-   AB_NYC_2019.csv ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐
-   Public School Characteristics ⭐⭐⭐⭐ 
-   USA Households ⭐⭐⭐⭐⭐

```{r}
cereal <- read_csv("_data/cereal.csv")
cereal
summary(cereal)

```

### Briefly describe the data

This dataset contains the data of sodium, sugar and type of different cereals in US. It is a small dataset with 20 rows and 4 columns.

## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

The Sodium and Sugar values are present in numbers but we don't know the units for it. The type column contains values A and C but we are not sure what they mean to mutate it thus we can leave it as it is. Thus there is not much we can do to tidy it,



```{r}
unique(cereal$Type)

```


## Univariate Visualizations

We can check the sugar and sodium levels across all cereals using a box plot. Thus will give us an overall of max, min, median quantity of sugar and sodium for cereals. I have also created a graph of top 5 cereals with highest sugar and sodium. Out of the 20 cereals, Raisin bran seems to have the highest sodium as well as highest sugar and thus should probably be avoided. 


```{r}
ggplot(cereal)+
  geom_boxplot(aes(x="Sugar", y=Sugar)) +
  labs(title = "Sugar level in Cereal", y="Quantity")

ggplot(cereal)+
  geom_boxplot(aes(x="Sodium", y=Sodium)) +
  labs(title = "Sodium level in Cereal", y="Quantity")

cereal_sugar <- head(arrange(cereal, desc(Sugar)),5)

ggplot(data = cereal_sugar, aes(x = Cereal, y = Sugar)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 5 High sugar cereals", x = "Cereal", y = "Sugar quantity")

cereal_sodium <- head(arrange(cereal, desc(Sodium)),5)

ggplot(data = cereal_sodium, aes(x = Cereal, y = Sodium)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 5 High Sodium cereals", x = "Cereal", y = "Sodium quantity")

```


## Bivariate Visualization(s)

I created a scatter plot of Sugar and Sodium for cereal which are distinguished by Type A or C based on the color. This graph gives us a picture of sodium and sugar level in cereal together.

```{r}
ggplot(cereal)+
  geom_point(aes(x = Sugar, y = Sodium, color = Type))+
  labs(title="Scatter plot of Sodium vs Sugar in Cereal boxes", x = "Sugar", y = "Sodium")
```