---
title: "Challenge 1: Reading and Understanding Data about Railroad Employees Across States & Counties"
author: "Hunter Major"
description: "Reading in data and creating a post"
date: "5/31/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - csv
  - railroads
  - total employees by state and county
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

For this challenge, I'll be reading in the following data set:

-   railroad_2012_clean_county.csv ⭐


###raiload_2012_clean_county.csv

#####Here's a snapshot of this dataset:
```{r}
read.csv("_data/railroad_2012_clean_county.csv")

read_csv("_data/railroad_2012_clean_county.csv")

total_employees_from_csv <- read_csv("_data/railroad_2012_clean_county.csv")

```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary
spec(total_employees_from_csv)

summary(total_employees_from_csv)

colnames(total_employees_from_csv)

dim(total_employees_from_csv)




```
###This dataset provides a snapshot into the United States' railroad workforce in 2012; the total number of railroad employees are sorted alongside their respective state and county.This data was likely compiled by either a governmental body, like the U.S. Bureau of Labor Statistics, or a railroad association, such as the Association of American Railroads. Historical societies, railroad museums, and websites like Ancestry.com may also have varying capacities to compile datasets similar to this. The dim () function details that this dataset has 2930 rows and 3 columns, making it a tall dataset as opposed to a wide one. Because there are 50 states in the United States-not including U.S. Territories, districts like D.C., APO/FPO addresses, etc, we can infer that this dataset sources information from 2930 counties across the country. The spec ()function lists the dataset's 3 columns as state-which is abbreviated, county, and total number of employees working for the railroad within a state/county. The colnames () function also provides insight into the names of the columns. The total_employees column has numerical variables, as it lists the number of railroad employees per county within a state. Whereas, the state and county columns have character values--abbreviating the names of states and listing the names of counties. Using the summary () function, we see that the number of railroad employees within any given county ranges from a minimum of one employee to a maximum of 8207 employees, which evidences that railroad employment varies considerably across the country/from county to county. Furthermore, the summary () function lists the mean as 87.18, meaning that on average, across the U.S., there are roughly 87 railroad employees per county. The median, listed in the summary () function as 21, is the middle value of the dataset, which means that most counties have railroad employee teams of a smaller quantity.