---
title: "Challenge 3"
author: "Noah Dixon"
description: "Tidy Data: Pivoting"
date: "6/08/2023"
format:
  html:
    df-print: paged
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_3
  - eggs
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in the Data

```{r}
#| label: read data
eggs_from_csv <- read_csv("_data/eggs_tidy.csv")
```

## Describe the Data

In order to get a better sense of what the data looks like, we can print the first 6 rows of the data using the head function.

```{r}
#| label: check data
head(eggs_from_csv)
```

It appears this data set is showing the number of eggs sold for different egg types during each month for specific years. Lets see how many years there are in the data set.

```{r}
#| label: check data again
print(unique(eggs_from_csv$year))
```

We have confirmed that the data set shows data for multiple years. For this data, it is clear that a case should be represented by a month, year, and container size (large_half_dozen, large_dozen, extra_large_half_dozen, and extra_large_dozen). Therefore, we need to pivot the data to remove the large_half_dozen, large_dozen, extra_large_half_dozen, and extra_large_dozen columns and add a container size column.


## Describe the Final Dimensions

The output below shows the dimensions of the current data fram in terms of number of rows and columns, as well as the expected dimensions of the new dataframe after pivoting.

```{r}
#existing rows/cases
nrow(eggs_from_csv)

#existing columns/cases
ncol(eggs_from_csv)

#expected rows/cases
nrow(eggs_from_csv) * 4

# expected columns 
ncol(eggs_from_csv) - 4 + 2

```

## Pivot the Data

Again, after pivoting a case should be represent a month, year, and container size.

```{r}
eggs_tidy <- pivot_longer(eggs_from_csv, col = large_half_dozen:extra_large_dozen, names_to = "container", values_to = "quantity")

eggs_tidy
```

As we can see, the pivoted data is the expected size (480 x 4).
