---
title: "Challenge 1 Solutions"
author: "Gabrielle Roman"
description: "Reading in data and creating a post"
date: "6/5/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - faostat
  - wildbirds
  - Gabrielle Roman
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
library(readr)
library(readxl)
library(dplyr)
list.files("_data")
railroad_from_csv <- read.csv("_data/railroad_2012_clean_county.csv")
railroad_from_csv
typeof(railroad_from_csv$total_employees)
typeof(railroad_from_csv$state)
typeof(railroad_from_csv$county)
typeof(railroad_from_csv$total_employees)
# spec(railroad_from_csv) isn't working on the basis that tbl_df is not true. I take it this means that my dataframe isn't a tibble.
railroad_from_csv %>% 
  select(state)%>%
  distinct()
head(railroad_from_csv)

```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary
# The data describes the quantity of employees by county and respective state in 2012. Variables include the state, county, and number of employees. 
```
