---
title: "Challenge 1 Solutions"
author: "Tenzin Latoe"
description: "Reading in data and creating a post"
date: "07/04/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - Tenzin Latoe
  - railroads
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

-   railroad_2012_clean_county.csv ⭐

```{r}
railroad <- read_csv("_data/railroad_2012_clean_county.csv")
```

```{r}
railroad
```


## Describe the data

The data consists of 2930 rows and 3 columns, which represent State, county, and total employees. 

```{r}
#Preview: railroad
head(railroad)
```
Head function used to preview the first 5 rows of the dataframe

```{r}
colnames(railroad)
```
```{r}
#Filter Massachusetts from railroad
filter(railroad, `state` == "MA")
```
To look specifically at MA, we can can filter the data by searching specific states. I selected MA from the current railroad dataset. 
