---
title: "Challenge 8"
author: "Prasann Desai"
description: "Joining Data"
date: "7/3/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - Prasann Desai
  - snl
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(readxl)
library(lubridate)
library(scales)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in multiple data sets, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  join two or more data sets and analyze some aspect of the joined data

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

  - military marriages ⭐⭐
  - faostat ⭐⭐
  - railroads  ⭐⭐⭐
  - fed_rate ⭐⭐⭐
  - debt ⭐⭐⭐
  - us_hh ⭐⭐⭐⭐
  - snl ⭐⭐⭐⭐⭐

```{r}
snl_seasons <- read_csv('_data/snl_seasons.csv')
snl_actors <- read_csv('_data/snl_actors.csv')
snl_casts <- read_csv('_data/snl_casts.csv')
```
## Previewing the datasets

```{r}
# snl_casts
nrow(snl_casts)
head(snl_casts)
```
```{r}
# snl_seasons
nrow(snl_seasons)
head(snl_seasons)
```
```{r}
# snl_actors
nrow(snl_actors)
head(snl_actors)
```

### Briefly describe the data

From the above outputs, we can see that the dataset contains 3 tables related to the popular show (SNL- Saturday Night Live).
- snl_actors contains a list of actors who appeared on SNL with a few characterstics about them like their url, gender and whether they were a cast/guest/crew/unknown member on SNL.
- snl_seasons contains a list of SNL seasons and basic information about it like first/last episode ids, num of episodes and year of the SNL season.
- snl_casts contains the details of actors that featured in every season of SNL and their stats like num of episodes they were part of, their first/last episode ids, etc. 


## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

Response: Overall, all the 3 tables look tidy.

Are there any variables that require mutation to be usable in your analysis stream? For example, do you need to calculate new values in order to graph them? Can string values be represented numerically? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

Document your work here.

```{r}
table(snl_actors$gender)
```

```{r}
# SNL actors dataset contains 'andy' gender. It would be a good idea to replace these instances with 'unknown'.
snl_actors <- mutate(snl_actors, gender = case_when(str_detect(gender, "andy") ~ "unknown",
                                                    str_equal(gender, "female") ~ "female",
                                                    str_equal(gender, "male") ~ "male",
                                                    TRUE ~ as.character(gender)
                                                    ))
# Sanity check of the fix
table(snl_actors$gender)
```

## Join Data

Be sure to include a sanity check, and double-check that case count is correct!

```{r}
# Joining the snl_casts and snl_actors on "aid" column
snl_ca <- left_join(snl_casts, snl_actors, by = "aid")
snl_ca
```

```{r}
# Joining the snl_ca and snl_seasons on "sid" column
snl_cas <- left_join(snl_ca, snl_seasons, by = "sid")
snl_cas
```

## Data Analysis

```{r}
snl_actors_years <- select(snl_cas, aid, year)
snl_yoy_actors <- count(snl_actors_years, year)
snl_yoy_actors <- mutate(snl_yoy_actors, 
                       year = as.Date(paste0(as.character(year), "0101"),format = "%Y%m%d"))
snl_yoy_actors
```

```{r}
# Line plot of trend of number of actors on SNL over years

#debt_pivoted$year_q <- yq(paste0(debt_pivoted$Year, ":", debt_pivoted$Quarter))

ggplot(snl_yoy_actors, aes(x = year, y = n)) +
  geom_point(size=0.5) +
  geom_line(size=0.5, color = "#FFDF00") +
  geom_smooth(alpha = 0.1, color = "#FFBF46") + 
  scale_x_date(date_breaks = '3 years',
               labels = date_format("%Y")) +
  scale_y_continuous(breaks = c(seq(1, 25, by=1))) +
  labs(title = "Number of actors on SNL (Y-o-Y)",
       subtitle = "1975-2020",
       x = "Years",
       y = "# Actors") +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2")
```