---
title: "Challenge 7 Solution"
author: "Shreya Varma"
description: "Visualizing Multiple Dimensions"
date: "7/9/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_7
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  Recreate at least two graphs from previous exercises, but introduce at least one additional dimension that you omitted before using ggplot functionality (color, shape, line, facet, etc) The goal is not to create unneeded [chart ink (Tufte)](https://www.edwardtufte.com/tufte/), but to concisely capture variation in additional dimensions that were collapsed in your earlier 2 or 3 dimensional graphs.
   - Explain why you choose the specific graph type
5) If you haven't tried in previous weeks, work this week to make your graphs "publication" ready with titles, captions, and pretty axis labels and other viewer-friendly features

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code. And anyone not familiar with Edward Tufte should check out his [fantastic books](https://www.edwardtufte.com/tufte/books_vdqi) and [courses on data visualizaton.](https://www.edwardtufte.com/tufte/courses)

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

  - eggs ⭐
  - abc_poll ⭐⭐
  - australian_marriage ⭐⭐
  - hotel_bookings ⭐⭐⭐
  - air_bnb  ⭐⭐⭐
  - us_hh ⭐⭐⭐⭐
  - faostat ⭐⭐⭐⭐⭐

```{r}

hotel_bookings <- read_csv("_data/hotel_bookings.csv")
head(hotel_bookings)
glimpse(hotel_bookings)
summary(hotel_bookings)

```

### Briefly describe the data

I will be using the Hotel Bookings dataset for my Challenge I have imported it using the read_csv() function and will use the glimpse() function the see the columns it has. On a high level it seems to have the information of hotel type for City Hotels and Resort Hotels and its customer data like arrival departure information, number of people, their booking details, payment type and reservation details. The data has 119,390 rows and 32 columns. To get more insights I will use the summary() function. From the summary we can see the data captured is from 2015 to 2017. 


## Tidy Data (as needed)

We can see that there are separate columns for arrival day of month, month and year. I will combine month and year to get a single monthly column for every year which is easier to read and I can plot the number of reservations by hotel and market segment date wise from 2015 to 2017. If I include day in this it will become too granular and the graph will be very congested and not readable. For this purpose I will change the month name to number and then mutate the two columns of month and year.

```{r}
hotel_bookings <- hotel_bookings %>%
                  mutate(arrival_date_month = case_when(
                    arrival_date_month == "January" ~ 1,
                    arrival_date_month == "Febuary" ~ 2,
                    arrival_date_month == "March" ~ 3,
                    arrival_date_month == "April" ~ 4,
                    arrival_date_month == "May" ~ 5,
                    arrival_date_month == "June" ~ 6,
                    arrival_date_month == "July" ~ 7,
                    arrival_date_month == "August" ~ 8,
                    arrival_date_month == "September" ~ 9,
                    arrival_date_month == "October" ~ 10,
                    arrival_date_month == "November" ~ 11,
                    arrival_date_month == "December" ~ 12
                  ))

hotel_bookings <- hotel_bookings %>%
                mutate(
                  arrival_date = make_date(arrival_date_year, arrival_date_month)
                ) 

hotel_bookings <- select(hotel_bookings,-c(arrival_date_month, arrival_date_year))

hotel_bookings

```

## Visualization with Multiple Dimensions

I created a graph of market segment based reservations over the months for City and Resort Hotels. We can see that Online TA remains to be the constant market segment with highest reservations. There is no change in market segment trend over the months and Online TA is followed by Offline TA/TO.

```{r}
ggplot(data = hotel_bookings, aes(x = arrival_date, fill = market_segment)) +
  geom_bar() +
  facet_wrap(vars(hotel))+
  labs(title = "Market segment based resevations over Time", x = "Month", y = "Number of reservations")
```