---
title: "Challenge 1"
description: "Reading in data and creating a post"
author: "Danny Holt"
date: "2023-06-01"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
execute:
  echo: false
---
```{r}
#| label: setup
#| warning: false

library(tidyverse)
library(readr)
knitr::opts_chunk$set(echo = TRUE)
```

## `birds.csv`
For this challenge, I will be reading in the dataset birds.csv
```{r}
birds <- readr::read_csv("_data/birds.csv")
```

Let's look at first several rows of the dataset:
```{r}
head(birds)
```
Year and YearCode appear to be duplicate variables.

As shown above, the data has 14 columns and 30977 rows. Let's look at the column names:
```{r}
colnames(birds)
```

Now, I will use `spec()` to inspect the data types of each of the columns in the dataset. Eight of the variables are categorical and six are numeric.
```{r}
spec(birds)
```

Here is a table of all of the types of birds found in the dataset under the column 'Item'. Chickens appear to be the most common type of bird here.
```{r}
table(birds$Item)
```

Now, we will use `colSums(is.na())` to see where data is missing. We see that some data is missing in the 'Value' and 'Flag' columns.
```{r}
colSums(is.na(birds))
```

The dataset counts different types of live birds (shown in column ‘Item’) in different areas (columns ‘Area’ and ‘Area Code’) and years (‘Year’ and ‘Year Code’). Based on the information in the ‘Flag Description’ column, the data appears to be a mix of collected data and estimates.