---
title: "Challenge 1 - Birds"
author: "Srujan Kagitala"
description: "Reading in data and creating a post"
date: "6/26/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - wildbirds
  - srujan_kagitala
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   birds.csv ⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
#| label: read
bird_data <- read.csv("_data/birds.csv")
str(bird_data) # To examine the structure of the bird_data object.
```

```{r}
#| label: preprocess
# Convert variables with character to factor format to get insights about data.
bird_data <- bird_data %>% mutate(Domain.Code = factor(Domain.Code),
                                  Domain = factor(Domain),
                                  Area = factor(Area),
                                  Element = factor(Element),
                                  Item = factor(Item),
                                  Unit = factor(Unit),
                                  Flag = factor(Flag),
                                  Flag.Description = factor(Flag.Description))
```


Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

The birds data set has `r nrow(bird_data)` observations and each observation has `r ncol(bird_data)` data points recorded. The variables recorded are `r names(bird_data)`. The below table shows a brief summary of each variable in the data set. For a quantitative variable, descriptive statistics are shown. Whereas, counts for unique values a variable can take are shown for categorical variables.

```{r}
#| label: summary
summary(bird_data)
```
Various types of birds like `r unique(bird_data$Item)` are considered in this data sample. This data set contains the collection source for each observation. These sources are `r unique(bird_data$Flag.Description)`. Below are pie chart and boxplot for variables Item and Value respectively.

```{r}
#| label: plots
pie(summary(bird_data$Item), col=rainbow(length(summary(bird_data$Item))), main="Types of Birds")
boxplot(bird_data$Value, horizontal=TRUE, main = "Boxplot for values", xlab = "values")
```