---
title: "Challenge 4"
author: "Prasann Desai"
description: "More data wrangling: pivoting"
date: "6/30/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_4
  - abc_poll
  - Prasann Desai
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  identify variables that need to be mutated
4)  mutate variables and sanity check all mutations

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   abc_poll.csv ⭐
-   poultry_tidy.xlsx or organiceggpoultry.xls⭐⭐
-   FedFundsRate.csv⭐⭐⭐
-   hotel_bookings.csv⭐⭐⭐⭐
-   debt_in_trillions.xlsx ⭐⭐⭐⭐⭐

```{r}
# Function call to read a csv file
abc_poll_2021 <- read_csv("_data/abc_poll_2021.csv")
```
```{r}
abc_poll_2021
```
```{r}
colnames(abc_poll_2021)
```


### Briefly describe the data

Response:

From the above output, we can see that the dataset contains certain demographic information about the 527 participants and their respective responses to all the questions of a poll conducted in 2021. Based on the column names, we can say that there were 10 different opinion/views based questions asked to each participant.

## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

Response: It appears to me that the data is already tidy. (Each variable has its own column, Each observation has its own row and Each value is in its own cell)

```{r}
# Sanity check for duplicate responses
n_distinct(abc_poll_2021$id)
```

```{r}
# Check to see if the category wise totals add up to 527
table(abc_poll_2021$ppeduc5)
```

```{r}
# Another Check to see if the category wise totals add up to 527
table(abc_poll_2021$ppethm)
```

Any additional comments?

## Identify variables that need to be mutated

Are there any variables that require mutation to be usable in your analysis stream? For example, are all time variables correctly coded as dates? Are all string variables reduced and cleaned to sensible categories? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

Document your work here.


```{r}
# Removing the redundant term "Non-Hispanic" from 'ppethm' column. Add a separate "Hispanic/Non-Hispanic" flag for backward compatibility.
abc_poll_2021 <- mutate(abc_poll_2021, is_hispanic = !str_detect(ppethm, "Non-Hispanic"), pprace = str_split(ppethm, ",", simplify = TRUE)[,1])
abc_poll_2021 %>% select(ppethm, is_hispanic, pprace)
table(abc_poll_2021$pprace)
```


```{r}
# Cleaning the 'ppeduc5' column values
abc_poll_2021 <- mutate(abc_poll_2021, ppeduc5_clean = str_replace(ppeduc5, "\x92", "'"))
abc_poll_2021 %>% select(ppeduc5_clean, ppeduc5)
table(abc_poll_2021$ppeduc5_clean)
```
```{r}
# Cleaning the 'contact' column to 'interview_consent'. Also, replacing the values to a simple "Yes" and "No" for better readability.
abc_poll_2021 <- rename(abc_poll_2021, interview_consent = Contact)
abc_poll_2021 <- mutate(abc_poll_2021, interview_consent_clean = case_when(str_detect(interview_consent, "Yes") ~ "Yes",
                                                                     str_detect(interview_consent, "No") ~ "No"
                                                                     ))
abc_poll_2021 %>% select(interview_consent, interview_consent_clean)
table(abc_poll_2021$interview_consent_clean)
```
Overall, I think the new columns added (is_hispanic, pprace, ppeduc5_clean, interview_consent_clean) will improve readability and understanding of the dataset as well as lead to cleaner visualizations that will be built on top of this dataset.


Any additional comments?
