---
title: "Challenge 9"
author: "Tenzin Latoe"
description: "Creating a function"
date: "07/18/2023"
format:
  html:
    df-print: paged
categories:
  - challenge_9
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is simple. Create a function, and use it to perform a data analysis / cleaning / visualization task:

Examples of such functions are:
1) A function that reads in and cleans a dataset.  
2) A function that computes summary statistics (e.g., computes the z score for a variable).  
3) A function that plots a histogram.  

That's it!

## Function 

```{r}
stats <- function (df) {
  stat <- c(
    Mean = mean(df, na.rm = TRUE),
    Median = median(df, na.rm = TRUE),
    Minimum = min(df, na.rm = TRUE),
    Maximum = max(df, na.rm = TRUE)
  )
  return(stat)
}
```
Created function to calculate mean, median, minimum, and maximum. 
```{r}
railroad <- read_csv("_data/railroad_2012_clean_county.csv")
```
Chose Railroad data set. 

```{r}
stats(railroad$total_employees)
```

Ran function. 
