---
title: "Challenge 6"
author: "Danny Holt"
description: "Visualizing Time and Relationships"
date: "6/20/2023"
format:
  html:
    df-print: paged
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - debt
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(readxl)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in data

First, let's read in the debt_in_trillions dataset.

```{r}
  #Read in data
  debt <- read_excel("_data/debt_in_trillions.xlsx")
  debt
```

### Briefly describe the data
The data shows quarterly measures (in trillions of dollars) of debt associated with six types of loans: `Mortgage`, `HE Revolving`, `Auto Loan` `Credit Card`, `Student Loan`, and `Other` (plus a `Total`). 

The data runs from the start of 2003 through the second quarter of 2021.

## Tidy Data

The data is mostly tidy, though we need to adjust the format in which dates are presented. We'll use `mutate()` and `parse_date_time()` to create a `Date` column with proper date formatting.

```{r}
# use parse_date_time to parse out dates from years and quarters
debt <- debt %>%
  mutate(Date=parse_date_time(`Year and Quarter`,
                              orders="yq"))
# remove `Year and Quarter` column
debt <- debt %>%
  select(-one_of("Year and Quarter")) %>%
  # move Date to front
  select(Date, everything())
# view data
debt
```

We're also going to create an alternate pivoted version of the data frame in which each observation contains data for a single type of debt with a date and amount. This version will have columns `Date`, `Type`, and `Amount`.
```{r}
debt_longer<- debt %>%
  pivot_longer(cols=c(`Mortgage`,`HE Revolving`,`Auto Loan`,`Credit Card`,`Student Loan`,`Other`), names_to = "Type", values_to = "Amount")
debt_longer<-debt_longer %>%
  select(Date,Type,Amount)
debt_longer
```

## Time Dependent Visualization
Now, let's look at total debt over time. We'll use a line plot (a variation of a scatter plot in which data points are connected with a line), which is good for visualizing change in one variable over time (a second variable).
```{r}
ggplot(debt, aes(x=Date,y=Total)) +
  geom_line() +
  labs(title = "Total US household debt", subtitle="2003 - 2021", x = "Date", y = "Debt (in trillions, USD)") +
  theme_minimal()
```

## Visualizing Part-Whole Relationships
Now, we'll look at debt over time, but broken out by type of debt. To do this, we'll use `debt_longer` from earlier. We'll use a stacked area chart, which is good for viewing changes in a set of things (in this case types of debt) where the cumulative value of all parts of the set is significant for analysis.
```{r}
# order types of debt with factor()
debt_longer$Type <- factor(debt_longer$Type, levels=c("Other","HE Revolving","Credit Card","Student Loan","Auto Loan","Mortgage"))
# create stacked area plot
ggplot(debt_longer, aes(x=Date,y=Amount,fill=Type)) +
  geom_area() +
  labs(title = "Total US household debt by type", subtitle="2003 - 2021", x = "Date", y = "Debt (in trillions, USD)") +
  theme_minimal()
```