---
title: "Challenge 3 Instructions"
author: "Danny Holt"
description: "Tidy Data: Pivoting"
date: "6/7/2023"
format:
  html:
    df-print: paged
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_3
  - eggs
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

2.  identify what needs to be done to tidy the current data
3.  anticipate the shape of pivoted data
4.  pivot the data into tidy format using `pivot_longer`

## Read in data

First, we'll read in the data set eggs_tidy.csv

```{r}
  #Read in data
  eggs <- readr::read_csv("_data/eggs_tidy.csv")
  #Number of rows and columns:
  nrow(eggs)
  ncol(eggs)
  #Data:
  eggs
```

### Briefly describe the data

The data set appears to show the number of sets of a dozen and half dozen large and extra large eggs, likely on a specific farm. Each observation/row shows data from a specific month. The data has 120 rows and 6 columns.

Because the columns other than `year` and `month` all measure number of eggs, we can pivot the data to combine the eggs together, and add the new columns `type` and `amount` to maintain the dozen/half dozen information and the amount information. Based on this, we know that the pivot will have four (2+2) columns.

We are effectively turning the existing columns `large_half_dozen`, `large_dozen`, `extra_large_half_dozen`, and `extra_large_dozen` into rows with the resulting rate of four rows in the pivot to one row in the existing data (four rows for a given month). This means that our data should have the following number of rows:
```{r}
nrow(eggs)*4
```

## Pivot the Data

Now we will pivot the data.

``` {r}
eggs<-
  eggs%>%
    pivot_longer(col=c(large_half_dozen,large_dozen,extra_large_half_dozen,extra_large_dozen),
                                 names_to="type",
                                 values_to="amount")
eggs
```
Here, we see that we do indeed have 480 rows and 4 columns.

A new "case" in the pivoted data will contain month and year data, as well as the type (dozen/half dozen and large/extra large) and amount.

In our pivoted data, each variable forms a column and each observation forms a row. All values have their own cells. The data is easier to examine because of this.
