---
title: "Homework 3: Diving into Data on Atlantic Hurricanes from 1920-2020"
author: "Hunter Major"
editor: source
description: Homework 3 Assignment on Atlantic Hurricanes Data
date: "6/29/2023"
format: 
 html:
    df-print: paged
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - hw3
  - atlantic_hurricanes
  - hunter_major
---

```{r}
# loading packages into R Studio


library(tidyverse)
library(dplyr)
library(readr)
library(readxl)
library(tidyr)
library(googlesheets4)
library(stringr)
library(lubridate)
library(ggplot2)


knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Homework 3 Overview
For this homework, your goal is to read in a more complicated dataset. Please use the category tag "hw2" as well as a tag for the dataset you choose to use.

Read in a dataset. It's strongly recommended that you choose a dataset you're considering using for the final project. If you decide to use one of the datasets we have provided, please use a challenging dataset - check with us if you are not sure. 

Clean the data as needed using dplyr and related tidyverse packages.
Provide a narrative about the data set (look it up if you aren't sure what you have got) and the variables in your dataset, including what type of data each variable is. The goal of this step is to communicate in a visually appealing way to non-experts - not to replicate r-code.
Identify potential research questions that your dataset can help answer.

```{r}
# reading in the dataset

atlantic_hurricanes <- 
read_csv("_mysampledatasets/atlantic_hurricanes.csv")


atlantic_hurricanes
```

## Narrative About The Data
```{r}
summary(atlantic_hurricanes)

dim(atlantic_hurricanes)
```

This dataset provides insights into Atlantic hurricanes, hurricanes that developed in the Atlantic Ocean area, across a 100 year time period, from 1920 to 2020. This dataset lists hurricanes that fall under the Category 1, Category 2, Category 3, and Category 5 classifications; therefore, it doesn't include Category 4 hurricanes nor does it include (storms that didn't develop beyond) tropical storms and tropical depressions. 458 hurricanes or observations/rows are included in this dataset. There are 10 variables. Variables in the original version of the data are: 

- ...1 or X (or the list number/ID of the hurricane as entered into the dataset, mostly used for organizational, data entry purposes)

- the name of the hurricane (character value), 

- the duration of the hurricane/the dates that it occurred (numeric/character value?),

- the wind speed of the hurricane (in miles per hour and kilometers per hour) (numeric value),

- the pressure of the hurricane (in atmospheric pressure-hPa and in inch of mercury-inHg) (numeric value),

- the number of deaths caused by the hurricane (numeric value),

- the amount of damage in US dollars caused by the hurricane (numeric value),

- the category of the hurricane (Cat 1, 2, 3, or 5) (number representing a character value?),

- the numerically assigned references/footnotes that provide further information about the hurricane (number representing a character value?.

**Using the summary () function, most of the values for each variable are listed as character values because they do include words, abbreviations, units of measurement, other symbols/punctuation in them currently, in/after the process of tidying the dataset, I hope to make sure that all of the variables that I've listed as having numeric values in paratheses above will be then 'understood'/interpreted by R as having numeric values. However, because I am a beginner in R--I don't know if I'll be able to accomplish all of what I perceive has to be cleaned within the suite of Homework 2 alone--as I will likely be working with this dataset (outside of the challenge assignments) all the way through to the final project, but I'll try my best:)

## Data Cleaning
#### 1. Removing REf Column and the ...1 or X Column
I am removing the REf column because I am unclear on what it represents, and I do not believe it will be useful for my purposes in tidying and working towards analyzing the snapshot this dataset provides into Atlantic hurricanes more broadly. I believe REf is potentially referring to listed/numbered footnote references in the original study. I sourced this dataset from Kaggle, https://www.kaggle.com/datasets/valery2042/hurricanes, and do not currently have access to the original study. The only information provided about sources on the dataset's Kaggle page is, "I scraped Wikipedia pages of Atlantic hurricanes of Categoris 1,2,3 and 5 using pandas/html" (Liamtsau, 2022). I am removing the ..1/X column as well because this is simply the list/ID number of each hurricane as it is entered into the dataset, and since R Studio maintains its own list/ID number on the far left of the table I believe the X variable is no longer necessary. Also, since the first number in the X column is 0 for the first hurricane listed instead of 1, this can be confusing for some readers whose numbering convention starts with 1. Scrolling all the way to the end of the table (page 46 for me), we can see that the last value listed in the X column is 457, which is a slight mismatch from the 458 rows/observation values, which represented the total number of hurricanes included in the study, that R computed the dataset to have.

```{r}
# remove column named REf and the X Column

atlantic_hurricanes2 <- atlantic_hurricanes %>% select(-c(...1,REf))
  

atlantic_hurricanes2 
```

Looks like the REf and X (or ...1) columns were successfully removed! There should now be 8 columns.

#### 2. Separate Wind.speed into Wind.speed.mph and Wind.speed.kmh and Pressure into Pressure.hPa and Pressure.inHg
In the current version of the dataset, within the Wind.speed column, values for each hurricane's wind speed are provided in miles per hour (mph) and kilometers per hour (km/h) in the same cell. Likewise, values for each hurricane's pressure are provided in hPa (atmospheric pressure) and inHg (inch of Mercury). I would like to separate those values, so each unit of measurement for the wind speed and pressure, respectively has their own distinct columns.

```{r}
# separate the Wind.speed column into Wind.speed.mph and Wind.speed.kmh

atlantic_hurricanes3 <- separate(atlantic_hurricanes2, `Wind speed`, into = c("Wind.speed.mph", "Wind.speed.kmh"), sep = "\\(")

atlantic_hurricanes3

```


```{r}
# separate Pressure column into Pressure.hPa and Pressure.inHg

atlantic_hurricanes4 <- separate(atlantic_hurricanes3, Pressure, into = c("Pressure.hPa", "Pressure.inHg"), sep = " ")

atlantic_hurricanes4

```
Looks like each unit of measurement for a hurricane's wind speed (Wind.speed.mph and Wind.speed.kmh) and a hurricane's pressure (Pressure.hPa and Pressure.inHg) now have their own distinct columns!

#### 3. Removing measurement unit abbreviations and unneeded parentheses from values in the Wind.speed.mph, Wind.speed.kmh, Pressure.hPa, and Pressure.inHg columns
I would like to remove the measurement unit abbreviations and unneeded parentheses from values in the Wind.speed.mph, Wind.speed.kmh, Pressure.hPa, and Pressure.inHg columns so that only the numbers/numeric values remain. Once R reads these columns as have numeric values, I'll be able to run summary statistics and other relevant numeric related functions using them that'll provide useful information to analyze.

```{r}
# removing "mph" from the end of values in the Wind.speed.mph column

atlantic_hurricanes5 <- mutate(atlantic_hurricanes4, Wind.speed.mph = as.numeric(str_extract(Wind.speed.mph,pattern="[:digit:]+")))

atlantic_hurricanes5

```

```{r}
# removing "km/h)" from the end of values in the Wind.speed.kmh column

atlantic_hurricanes6 <- mutate(atlantic_hurricanes5, Wind.speed.kmh = as.numeric(str_extract(Wind.speed.kmh,pattern = "[:digit:]+")))

atlantic_hurricanes6

```

```{r}
# removing commas from values in Pressure.hPa
# removing "hPa" from the end of values in the Pressure.hPa column

atlantic_hurricanes7<- mutate(atlantic_hurricanes6, Pressure.hPa = str_remove(Pressure.hPa, ","),
                              Pressure.hPa= as.numeric(str_extract(Pressure.hPa,pattern = "[:digit:]+")))
atlantic_hurricanes7



```

```{r} 
# removing "(" and "inHg)" from values in Pressure.inHg column

#areas affected
n_areas_max <- max(str_count(atlantic_hurricanes7[!is.na(atlantic_hurricanes7$`Areas affected`),]$`Areas affected`, "[a-z],"))+2

#separate areas affected into multiple columns then pivot longer into one column, called area
atlantic_hurricanes8 <- atlantic_hurricanes7 %>%
  separate(`Areas affected`, into = paste0("a",1:n_areas_max), sep = ",") %>%
  pivot_longer(c(a1,a2,a3,a4,a5,a6,a7,a8,a9,a10,a11),names_to = "del", values_to= "area") %>%
  select(-del)%>%
  filter(!is.na(area))
atlantic_hurricanes8

write_csv(atlantic_hurricanes8, "atlantic_hurricanes8.csv")


```

## Reading Back in the Data Set after manually changing small things in Google Sheets
I made a few minor changes to the atlantic_hurricanes8 dataset in Google Sheets such as removing cross signs (a special character) from the end of some duration dates, changing the dashes in the duration column from double dash (--, a special character) to a single dash (-), duplicating rows to separate two or more different affected areas that previously were not listed with spaces between them and changing the name of the Category column to Max.category.
```{r}
# reading back in data set after manually changing some small things in Google Sheets
atlantic_hurricanes9 <- read_csv("_mysampledatasets/atlantic_hurricanes8_GoogleSheetsVersion.csv")
atlantic_hurricanes9

```

## Additional Cleaning
### Removing Wind.speed.kmh and Pressure.inHg columns and renaming Wind.speed.mph to Max.wind.speed.mph and Pressure.hPa into Max.pressure.hPa
```{r}
# Deleting Wind.speed.kmh and Pressure.inHg columns so that there's only one measure for wind speed (mph) and one measure for pressure (hPa)

atlantic_hurricanes10 <- atlantic_hurricanes9 %>%
  select(-c(Wind.speed.kmh, Pressure.inHg))
atlantic_hurricanes10

atlantic_hurricanes11 <- atlantic_hurricanes10 %>%
  mutate(Start_Date=str_c(str_extract(Duration,"[:alpha:]+ [:digit:]+(?=,)"),
      str_extract(Duration,", [:digit:]+")))
atlantic_hurricanes11

# renaming Wind.speed.mph to Max.wind.speed.mph and Pressure.hPa to Max.pressure.hPa
atlantic_hurricanes12 <- atlantic_hurricanes11 %>%
  rename(Max.wind.speed.mph=Wind.speed.mph)%>%
  rename(Max.pressure.hPa=Pressure.hPa)
atlantic_hurricanes12

```
Looks like Wind.speed.kmh and Pressure.inHg were removed successfully!

### Tidying the Deaths column: changing "None" values to 0 and "Unknown" values to NA and changing Deaths column to read as a numeric value
```{r}
# change values that read in Deaths column as "None" to the number 0
atlantic_hurricanes12$Deaths <- str_replace(atlantic_hurricanes12$Deaths, "None", "0")
atlantic_hurricanes12

# change values in Deaths column that read as "Unknown" to NA
atlantic_hurricanes12$Deaths <- na_if(atlantic_hurricanes12$Deaths, "Unknown")
atlantic_hurricanes12

# change values in Deaths column that contain >1,000 to NA
atlantic_hurricanes12$Deaths <- na_if(atlantic_hurricanes12$Deaths, ">1,000")
atlantic_hurricanes12

# change Deaths column to read as a numeric variable
atlantic_hurricanes13 <- transform(atlantic_hurricanes12, Deaths = as.numeric(Deaths))
atlantic_hurricanes13
```
As expected, looks like the Deaths column now reads as a numeric variable and the "None" value has been switched to 0 and the "Unknown" value has been switched to NA!

### Tidying the Damage column
Currently, there are a handful of values in the Damage column that will not make for the clearest analysis. All of the values contained in the Damage column can be seen when running the unique () function, which I will do below. I will change values that do not make for the clearest analysis/are less straightforward to NA. I will also remove the dollar signs, assuming that the creator of this dataset used USD for all monetary values. This is part of the process of having Damage eventually read as a numeric variable--so I can compute summary statistics and do visualizations off of the dollar amount. Like the Deaths column, there is also a value in the Damage column called "None" but instead of changing that to 0 I will be changing that to NA because I find it hard to believe that a hurricane (even of a lower intensity) caused zero damage (as a qualitative descriptor) and/or $0 worth of damage as a more quantitative descriptor.

```{r}
# checking for all unique values in the Damage column
unique(atlantic_hurricanes13$Damage)
```

```{r}
# changing less clear/straightforward values to NA

atlantic_hurricanes13$Damage[startsWith(atlantic_hurricanes13$Damage, ">")] <- NA

atlantic_hurricanes13$Damage[startsWith(atlantic_hurricanes13$Damage, "[")] <- NA

atlantic_hurricanes13$Damage[startsWith(atlantic_hurricanes13$Damage, "M")] <- NA

atlantic_hurricanes13$Damage[startsWith(atlantic_hurricanes13$Damage, "H")] <- NA

atlantic_hurricanes13$Damage[startsWith(atlantic_hurricanes13$Damage, "U")] <- NA

atlantic_hurricanes13$Damage[startsWith(atlantic_hurricanes13$Damage, "None")] <- NA

atlantic_hurricanes13$Damage[startsWith(atlantic_hurricanes13$Damage, "E")] <- NA

atlantic_hurricanes13$Damage[startsWith(atlantic_hurricanes13$Damage, "13")] <- NA

atlantic_hurricanes13$Damage[startsWith(atlantic_hurricanes13$Damage, "5")] <- NA
```

```{r}
atlantic_hurricanes14 <- atlantic_hurricanes13
atlantic_hurricanes14

unique(atlantic_hurricanes14$Damage)
```
Looks like all of the less clear/less straightforward values in the Damage column have been removed!

```{r}
# separating Damage into Damage.amount and Damage.unit
atlantic_hurricanes15 <- atlantic_hurricanes14%>%
  separate(Damage, c("Damage.$.amount", "Damage.$.unit"), " ")
atlantic_hurricanes15

# removing $ dollar sign from Damage.$.amount column, making Damage.$.amount column a numeric variable
atlantic_hurricanes15$`Damage.$.amount`= as.numeric(gsub("\\$", "", atlantic_hurricanes15$`Damage.$.amount`))
atlantic_hurricanes15

```

Looks like the Damage column has been separated into two different columns, and the Damage.$.amount column is now a numeric variable!

### Revisiting/Further Cleaning the area column
I'd like to change "No land areas" and "None" values within the area column to NA; with that said, in order to make that change, I think it's fitting to rename the area column to Land.areas.affected, so NA in that case could mean the land areas affected were/are unknown or there were no land areas affected, as in the hurricane system only remained in open waters and did not formally make landfall at any land-based location/territory.
```{r}
# Renaming area to Land.areas.affected
atlantic_hurricanes16 <- atlantic_hurricanes15 %>%
  rename(Land.areas.affected=area)
atlantic_hurricanes16

# viewing all unique values in Land.areas.affected
unique(atlantic_hurricanes16$Land.areas.affected)

# changing "None" and "No land areas" and "No Land Areas" to NA
atlantic_hurricanes16$Land.areas.affected[startsWith(atlantic_hurricanes16$Land.areas.affected, "None")] <- NA

atlantic_hurricanes16$Land.areas.affected[startsWith(atlantic_hurricanes16$Land.areas.affected, "No land areas")] <- NA

atlantic_hurricanes16$Land.areas.affected[startsWith(atlantic_hurricanes16$Land.areas.affected, "No Land Areas")] <- NA

#re-checking for NA in Land.areas.affected column
atlantic_hurricanes17 <- atlantic_hurricanes16
atlantic_hurricanes17

unique(atlantic_hurricanes17$Land.areas.affected)

```
Looks like modifications to the Land.areas.affected column were made successfully as expected!

## Summary Statistics
### summary stats about Max.wind.speed.mph
```{r}
# summary stats about Max.wind.speed.mph
summary(atlantic_hurricanes17$Max.wind.speed.mph)
```
### summary stats about Max.pressure.hPa
```{r}
#summary stats about Max.pressure.hPa
summary(atlantic_hurricanes17$Max.pressure.hPa)
```
### summary stats about Deaths
```{r}
#summary stats about Deaths
summary(atlantic_hurricanes17$Deaths)
```

### summary stats about Land.areas.affected
```{r}
# summary stats about total number of distinct land areas affected
atlantic_hurricanes17%>% summarise(count = n_distinct(Land.areas.affected))

# summary of how many times each land area was featured
table(atlantic_hurricanes17$Land.areas.affected)
```

### summary stats about Max.category
```{r}
# summary stats about total number of distinct Max categories featured
atlantic_hurricanes17%>% summarize(count = n_distinct(Max.category))

# summary stats about how many times each distinct Max category is featured
table(atlantic_hurricanes17$Max.category)
```
### summary stats about Name
```{r}
#summary stats about total number of distinct hurricane names featured
atlantic_hurricanes17%>% summarize(count = n_distinct(Name))

#summary stats about how many times each hurricane name was featured-->there are repeats because of how we pivoted land areas affected earlier but also hurricane names are periodically recycled
table(atlantic_hurricanes17$Name)
```

### summary stats about Damage.$.units

```{r}
# summary stats number of distinct Damage.$.units
atlantic_hurricanes17%>% summarise(count = n_distinct(`Damage.$.unit`))

# summary stats of each Damage.$.units featured-->should show that most of the hurricanes cost in the millions range
table(atlantic_hurricanes17$`Damage.$.unit`)

```
## Note: Fully Tidying both the Duration column and the Damage.$.amount column is still in progress! *this and is relevant summary stats and visualizations incorporating them will be fully updated by the final project due date

## Data Visualizations
### making temporary modifications to duration column and pairing down the number of years from 1920-2020 down to 2000-2020 for visualization purposes in this draft
```{r}
# Modifying the Duration column into Date and Year
atlantic_hurricanes18 <- atlantic_hurricanes17 %>%
  separate(Duration, c("Date", "Year"), ", ")
atlantic_hurricanes18

# Editing down to only focus on years 2000-2020
atlantic_hurricanes19 <- atlantic_hurricanes18[atlantic_hurricanes18$Year >= "2000" & atlantic_hurricanes18$Year <= "2020",]
atlantic_hurricanes19

```

## Connected Scatterplot of Correlation between Max.wind.speed.mph and Max.pressure.hPa
```{r}
# Scatterplot of Correlation between Max.wind.speed.mph and Max.pressure.hPa
ggplot(atlantic_hurricanes19, aes(x= Max.wind.speed.mph, y= Max.pressure.hPa))+
  geom_point(size=3, fill="black", color="black")+
  geom_line(color= "red", size= 1)
  labs(title = "Connected Scatterplot of Correlation between Max.wind.speed.mph & Max.pressure.hPa", subtitle = "2000-2020", x= " Max Wind Speed (mph)", y="Max Pressure (hPa)")+
  theme(axis.text.x = element_text(angle = 30, size = 2))+
  facet_grid()+
  theme_minimal()

```
For the most part, it seems as though as Wind Speed (mph) increases, Pressure (hPa) decreases--meaning that more severe hurricanes (of a higher category) have lower pressure (hPa) measures.

## Bar Plot of Max Category by Max.wind.speed.mph sorted by Hurricane Name, 2018-2020
```{r}
#pair down to storms between 2018-2020

atlantic_hurricanes20 <- atlantic_hurricanes19[atlantic_hurricanes19$Year >= "2018" & atlantic_hurricanes19$Year <= "2020",]
atlantic_hurricanes20

# bar plot of Name by Max.wind.speed.mph
ggplot(atlantic_hurricanes20, aes(x= Max.category, y= Max.wind.speed.mph, fill= Name))+
  geom_bar(stat = "identity", position= position_dodge(width = .6), width = 0.4)+
  labs(title = "Bar Plot of Max Category by Max Wind Speed sorted by Hurricane Name", subtitle= "2018-2020, **Category 4 data not included in dataset", x= "Max Category (no storms w/ max cat. 1 in these years)", y="Max Wind Speed (mph)", fill= "Hurricane Name")+
  facet_grid()+
  theme_light()

```
Looks like half of the hurricanes (6 out of 12) from 2018-2020 were Category 2 storms with max wind speeds hovering slightly over 100 mph.

## Bar Plot of Hurricane Frequency over the Years, 2000-2020 sorted by Damage.$.unit
```{r}
# Bar Plot of Hurricane Count over the Years, 2000-2020, sorted by Damage.$.unit
ggplot(data = atlantic_hurricanes19, aes(x= Year, fill= `Damage.$.unit`))+
  geom_bar(width= .2, position= position_dodge(width = .4))+
  facet_grid()+
  theme(axis.text.x = element_text(angle = 30, size = 7))+
  labs(title= "Bar Plot of Hurricane Count over the Years sorted by unit of Damage Cost ($USD)", subtitle= "2000-2020", x= "Year", y= "Count", fill= "Damage.$.unit")
  
```
Overall, notwithstanding the damage $ unit values that are NA or non-available, it seems like the damage from hurricanes across 2000-2020 is becoming more expensive--even in more recent years where it may seem like there's slightly less hurricanes per year (compared to a year like 2005), more hurricanes are more consistently causing damage in the billions of dollars...the last recorded hurricanes where the damage cost stayed within the thousands unit was in the year 2006--over 15 years ago.

## Potential Research Questions This Dataset Could Help Answer
Do duration, wind speed, pressure, and hurricane category classification have any correlation with deaths and damage costs caused by hurricanes? If so, how so? Which areas are most susceptible to the most severe and devastating hurricanes in the century this dataset covers? On the other hand, there are some limitations to this dataset. Something I noticed this dataset doesn't answer is, at what point in the duration of the hurricane were these measures of wind speed and pressure taken? While I'm assuming it is a listing of the highest intensity recorded wind speed and pressure throughout the course of the storm, this isn't clear. Moreover, for the hurricanes that impacted multiple areas, the windspeed and pressure was likely not the exact same measure in each successive location the hurricane made landfall in. While insights can be deduced from this dataset, especially after additional, forthcoming stages of cleaning the data, studying each storm and each area individually will provide additional, more full and nuanced context(s) of systems, resources, lived experiences, etc that influenced and/or were influenced by the hurricane(s).

## References
Liamtsau, V. (2022, August 5). Atlantic hurricanes (data cleaning challenge). Kaggle. https://www.kaggle.com/datasets/valery2042/hurricanes