---
title: "Challenge 6 Instructions"
author: "Sean Conway"
description: "Visualizing Time and Relationships"
date: "5/30/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - fed_rate
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in data

The data that I chose to read in was the `RedFundsRate.csv`

  - fed_rate ⭐⭐


```{r}

fed_rate <- read_csv("_data/FedFundsRate.csv")

fed_rate

```

### Briefly describe the data

## Tidy Data (as needed)

The only variable I needed to mutate was the `Year`, `Month` and `Day` into a single date object. I first combined them into one column, then used the `as.Date` function.

```{r}
date_col <- paste(fed_rate$Year, fed_rate$Month, fed_rate$Day)

date_col <- as.Date(date_col, format = "%Y %m %d")

head(date_col)

fed_rate <- fed_rate[-c(1, 2, 3)]

fed_rate <- fed_rate %>% add_column(`Date`=date_col, .after=0)

head(fed_rate)

```

The variables that I mutated were most of the rates. This included the `Effective Fed Funds Rate`, `Unemplyoment Rate`, and `Inflation Rate`. I mutated these to allow us to compare how they changed through the years and graphing them a bit simpler.

These new variables were created by the `diff` command, and can be seen in the last 3 columns.

```{r}

effective_fed_fund_change <- diff(fed_rate$`Effective Federal Funds Rate`)
effective_fed_fund_change <- append(effective_fed_fund_change, NA, after=0)

fed_rate['Effective Fed Funds Rate Change'] <- effective_fed_fund_change

unemployment_change <- diff(fed_rate$`Unemployment Rate`)
unemployment_change <- append(unemployment_change, NA, after=0)

fed_rate['Unemployment Rate Change'] <- unemployment_change

inflation_rate_change <- diff(fed_rate$`Inflation Rate`)
inflation_rate_change <- append(inflation_rate_change, NA, after=0)

fed_rate['Inflation Rate Change'] <- inflation_rate_change

head(fed_rate)

```

## Time Dependent Visualization

For the time dependent I chose to graph the inflation rate.

```{r}

ggplot(fed_rate, aes(x=`Date`, y=`Inflation Rate`)) +
  geom_line()

```

I chose to do a line graph because I believe it is the best way to graph a time-dependent variable. It offers a good visualization of how the inflation rate from year to year is fairly constant, aside from a few hiccups.

I chose to not include the points as I thought they made the graph too noisey and took away from the finer details.

## Visualizing Part-Whole Relationships

For the Part-Whole Relationship Visualization, I chose to graph inflation rate change over the samples given.

```{r}

#rates <- fed_rate %>% select(`Effective Federal Funds Rate`, `Unemployment Rate`, `Inflation Rate`)

#fed_rate$`Inflation Rate Change`

ggplot(fed_rate, aes(x=`Date`, y=`Inflation Rate Change`)) +
  geom_line()


```
I chose a line graph because it made the most logical sense on how to graph a variable where we are trying to see the change over a period of time. It also allows the viewer to see a nice progression of how the inflation rate change stays fairly constant over the years.

Similarly I omitted the dots because they were too large for how many points I had.